Double White Dwarf Mergers on Adaptive Meshes\\ I. Methodology and Code Verification

### 2015-07-31 papers/wdmerger_I/paper.tex
Define the projection operator in the strain tensor definition, from BDS.

We follow the prescription of \citet{blanchet:1990}. At distances far from the 
gravitational wave source, we can consider the leading term in the gravitational 
wave signal:
\begin{equation}
  h^{TT}_{ij}(t,\mathbf{x}) = \frac{2G}{c^4 r}P_{ijkl}(\mathbf{n}) \ddot{Q}_{kl}(t - r/c).
\end{equation}
$h$ is the perturbation to the spacetime metric and is commonly called the \textit{strain}; 
for laser interferometers, it measures the relative change in the distance between mirrors. 
The ``TT'' superscript indicates that we work in the commonly used tranverse-traceless gauge.
This strain is measured at time $t$ and position $\mathbf{x}$ relative to the binary system.
$r\equiv |\mathbf{x}|$ is the distance from the observer to the binary system. The unit vector 
$\mathbf{n} \equiv \mathbf{x} / r$ then measures the direction of the outgoing wave with 
respect to the observer, and $P_{ikjl}(\mathbf{n})$ is an operator in the transverse-traceless 
gauge that projects a tensor onto the direction orthogonal to $\mathbf{n}$. $Q_{kl}$ is the 
quadrupole moment tensor:
\begin{equation}
  Q_{kl} = \int dV \rho \left(x^k x^l - \frac{1}{3}\delta^{kl} \mathbf{x}^2\right).
\end{equation}
The argument $(t - r/c)$ indicates that to get the strain at time $t$ we evaluate the second derivative of the 
quadrupole moment at the retarded time $t - r/c$. In practice we ignore this in calculating this strain,
instead using the current simulation time, and so this merely yields a time delay of order $r/c$ between 
when the waves are emitted and when we observe them.

@@

The strategy is then as follows. At the end of the coarse timestep, we first calculate $\ddot{Q}_{kl}$
using an integral over the domain (using refined grids as usual by masking out the contributions from 
coarse grids underlying more accurate fine grids). This quantity is independent of the observer. If we 
are using a rotating reference frame, we first convert velocities and positions back to the inertial 
frame before evaluating the integral. Then, 
we pick an observing location $\mathbf{x}$ relative to the domain, evaluate the projection operator, 
and then perform the relevant tensor contraction to determine the strain tensor. We can 
repeat this process for any number of observing locations at minimal cost, since the quadruple tensor 
only needs to be calculated once. Now, gravitational waves only excite modes orthogonal to their 
direction of travel. These are the ``plus'' and ``cross'' modes, $h_+$ and $h_\times$, named after 
the types of spatial distortions they exhibit. We calculate the signal at a distance $r$ along 
the $x$, $y$ and $z$ axes. For the latter, as an example, $h_{+} = h_{11} = -h_{22}$ and 
$h_{\times} = h_{12} = h_{21}$. All other entries vanish. By default we take $r = 10$ kpc; 
as shown by \citet{loren-aguilar:2005}, this is a typical distance scale over which an 
experiment such as LISA could detect a coalescing binary white dwarf system. 
However, the strain at any other distance is easily calculated and goes as the inverse of the distance.

@@

The software used to generate the test problems in this paper
(as well as the manuscript itself),
\texttt{wdmerger}\footnote{\url{https://github.com/BoxLib-Codes/wdmerger}},
is freely available at an online repository hosting service.
Version control in both the parent software (BoxLib, CASTRO) and in \texttt{wdmerger}
permits us to reference the state of the code at the time a simulation
was performed. In all plot files and diagnostic output generated by CASTRO, 
and EPS figure files and plots generated by \texttt{wdmerger} for this work,
we store the active \texttt{git} commit hashes of BoxLib, CASTRO and \texttt{wdmerger}.
Line plots are generated using the \texttt{matplotlib} library for \texttt{Python} 
\cite{matplotlib}, while slice plots and other multi-dimensional visualizations are 
generated using the \cite{yt} code.

>>>>

We follow the prescription of \citet{blanchet:1990}. At distances far from the 
gravitational wave source, we can consider the leading term in the gravitational 
wave signal:
\begin{equation}
  h^{TT}_{ij}(t,\mathbf{x}) = \frac{2G}{c^4 r}P_{ijkl}(\mathbf{n}) \ddot{Q}_{kl}(t - r/c).
\end{equation}
$h$ is the perturbation to the spacetime metric and is commonly called the \textit{strain}; 
for laser interferometers, it measures the relative change in the distance between mirrors. 
The ``TT'' superscript indicates that we work in the commonly used tranverse-traceless gauge.
This strain is measured at time $t$ and position $\mathbf{x}$ relative to the binary system.
$r\equiv |\mathbf{x}|$ is the distance from the observer to the binary system. The unit vector 
$\mathbf{n} \equiv \mathbf{x} / r$ then measures the direction of the outgoing wave with 
respect to the observer, and $P_{ikjl}(\mathbf{n})$ is an operator in the transverse-traceless 
gauge that projects a tensor onto the direction orthogonal to $\mathbf{n}$:
\begin{align}
  P_{ijkl}(\mathbf{n}) &= \left(\delta_{ik} - n_i n_k\right)\left(\delta_{jl} - n_j n_l\right) \notag \\
                      &- \frac{1}{2}\left(\delta_{ij} - n_i n_j\right)\left( \delta_{kl} - n_k n_l\right).
\end{align}
$Q_{kl}$ is the quadrupole moment tensor:
\begin{equation}
  Q_{kl} = \int dV \rho \left(x^k x^l - \frac{1}{3}\delta^{kl} \mathbf{x}^2\right).
\end{equation}
The argument $(t - r/c)$ indicates that to get the strain at time $t$ we evaluate the second derivative of the 
quadrupole moment at the retarded time $t - r/c$. In practice we ignore this in calculating this strain,
instead using the current simulation time, and so this merely yields a time delay of order $r/c$ between 
when the waves are emitted and when we observe them.

@@

The strategy is then as follows. At the end of the coarse timestep, we first calculate $\ddot{Q}_{kl}$
using an integral over the domain (using refined grids as usual by masking out the contributions from 
coarse grids underlying more accurate fine grids). This quantity is independent of the observer. If we 
are using a rotating reference frame, we first convert velocities and positions back to the inertial 
frame before evaluating the integral. Then, 
we pick an observing location $\mathbf{x}$ relative to the domain, evaluate the projection operator, 
and then perform the relevant tensor contraction to determine the strain tensor. We can 
repeat this process for any number of observing locations at minimal cost, since the quadruple tensor 
only needs to be calculated once. Now, gravitational waves only excite modes orthogonal to their 
direction of travel. These are the ``plus'' and ``cross'' modes, $h_+$ and $h_\times$, named after 
the types of spatial distortions they exhibit. We calculate the signal at a distance $r$ along 
the $x$, $y$ and $z$ axes. For the latter, as an example, $h_{+} = h_{11} = -h_{22} \propto (\ddot{Q}_{11} - \ddot{Q}_{22})/2$ and 
$h_{\times} = h_{12} = h_{21} \propto \ddot{Q}_{12}$. All other entries vanish. By default we take $r = 10$ kpc; 
as shown by \citet{loren-aguilar:2005}, this is a typical distance scale over which an 
experiment such as LISA could detect a coalescing binary white dwarf system. 
However, the strain at any other distance is easily calculated and goes as the inverse of the distance.

@@

The software used to generate the test problems in this paper
(as well as the manuscript itself),
\texttt{wdmerger}\footnote{\url{https://github.com/BoxLib-Codes/wdmerger}},
is freely available at an online repository hosting service.
Version control in both the parent software (BoxLib, CASTRO) and in \texttt{wdmerger}
permits us to reference the state of the code at the time a simulation
was performed. In all plot files and diagnostic output generated by CASTRO, 
and EPS figure files and plots generated by \texttt{wdmerger} for this work,
we store the active \texttt{git} commit hashes of BoxLib, CASTRO and \texttt{wdmerger}.
Line plots are generated using the \texttt{matplotlib} library for \texttt{Python} 
\citep{matplotlib}, while slice plots and other multi-dimensional visualizations are 
generated using the \texttt{yt} code \citep{yt}.
